%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}%

\begin{ntacknowledgements}

First, I would like to thank Carla Ferreira for being my adviser since my third
  year. The knowledge and guidance I have received from her are truly
  invaluable. I am grateful for all the tips, her patience when things did not
  go as planned, her constant availability, and for being the incredible mentor
  that she is. She is the most professional and inspiring teacher I have had
  the pleasure to learn from and work with, and for that, I am immensely
  thankful.

To Bernardo Toninho, thank you for all your guidance, encouragement, and the
  freedom you gave me to explore my ideas and discover new paths, while always
  being there to support me. The lessons I learned and the discoveries I made
  under your mentorship are priceless. Working with you and experiencing your
  excellence as a teacher is something I will always cherish.

Thank you both for being my advisers, the best mentors I could have asked
  for. 

  I would also like to thank everyone that has been part of this journey, either
  simply watching me or living it with me: to my mother, Svitlana, for her
  unconditional support and for always rooting for me, even when things did not
  go according to plan, without her I would not be able to achieve half of the
  things I achieved; to my father, Mykola, for his important lessons about life
  and for never doubting me, even when he did not understand what I was doing
  most of the time; to my sister, Mariana, for \sout{annoying me} supporting me and listening
  to my complaints when things got south; to my dog Tracy, for all the support
  sitting on my lap on late night coding sessions. To all of my friends, that
  always offered me invaluable support, company, and friendship. Without them
  this would not have been as fun as it was.

  I would also like to dedicate this work to the heroes of my homeland, Ukraine, for fighting
  for freedom and life. Slava Ukraini!

  Thank you all so much! I am forever grateful.

\end{ntacknowledgements}
